### 2018-04-10 latex/egsr17-cgf/section_discretized_rte.tex
final version

Our method is derived from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
}\nonumber\\
&
+Q\left(\vec{x}, \omega\right)\nonumber
\label{eq:rte}
\end{align}

The left hand side (LHS) is the transport term and we refer to the terms on the right hand side (RHS) as collision, scattering and source term respectively. The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.

The RTE is often given in operator notation, where transport, collision and scattering are expressed as operators $\mathcal{T}$, $\mathcal{C}$ and $\mathcal{S}$, which are applied to the radiance field $L$:
\begin{align}
\mathcal{T}\left(L\right) = -\mathcal{C}\left(L\right) + \mathcal{S}\left(L\right) + Q
\end{align}

Deterministic methods, like ours, are derived by discretizing the angular and spatial domain. This gives rise to a linear system of equations, which can be solved using standard methods. For the $P_N$-method, the angular variable is discretized first, using a truncated spherical harmonics expansion. This results in the $P_N$-equations, a system of coupled PDE's, which still have a continuous spatial variable.

The number of equations grows with the truncation order $N$. This is why the discretization is hard and error prone if done by hand. We therefore use a computer algebra representation to automate this process. After giving an outline of the general discretization in this section, we will present our computer algebra representation in the next section. The $P_N$-equations, as result of our automated discretization, are given in section~\ref{sec:real_valued_pn_eq}.

Since the radiance field $L$ is real only, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
\begin{align}
\SHBR^{l,m}=
\left\{
\begin{array}{lr}
\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
\SHBC^{l,m}, & \text{for } m = 0\\
\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
\end{array}
\right.
\label{eq:sh_real_basis}
\end{align}

We express the projection into the SH basis functions with the projection operator $\mathcal{P}$:
\begin{align}
\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\mathbf{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}
\nonumber
\end{align}

The $P_N$-equations are derived by first expressing all angular dependent parameters in spherical harmonics. The radiance field $L$ in equation~\ref{eq:rte} is therefore replaced by its SH reconstruction $\widehat{L}$, introducing an error due to truncation at order $N$:
\begin{align}
\widehat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
}
\approx
L\left(\vec{x}, \omega\right)
\nonumber
\end{align}

After substitution, all angular parameters are expressed in terms of spherical harmonics, but still depend on the continuous angular variable $\omega$. As a next step, we project each term of the RTE into SH, using the projection operator $\mathcal{P}$. This produces a single equation for each $l,m$-pair. The $P_N$-equations therefore can be written as:
\begin{align}
\mathcal{P}^{l,m}\mathcal{T}\left(\widehat{L}\right)
=
-\mathcal{P}^{l,m}\mathcal{C}\left(\widehat{L}\right) 
+\mathcal{P}^{l,m}\mathcal{S}\left(\widehat{L}\right)
+\mathcal{P}^{l,m}\left(Q\right)
\label{eq:pn_operator_notation}
\end{align}

Once the $P_N$-equation have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxelgrid (using central differences for differential operators).

The radiance field $L$, after full discretization, is represented as a set of SH coefficients per voxel. Flattening these over all voxels into a single vector, gives the solution vector $\vec{u}$. The RHS vector $\vec{Q}$ is produced similarly. The projected operators can be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$ (see figure~\ref{fig:matrix_layout}):
\begin{align}
(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
\end{align}

@@

So far we have given the $P_N$-equations only in high-level operator notation (equation~\ref{eq:pn_operator_notation}). Carrying out the full derivation creates large, unwieldy equations and requires a string of expansions, manipulations and applications of identities. These are hard to do and error prone if done by hand. We therefore developed a computer algebra representation, which allowed us to derive the the $P_N$-equation in a semi-automatic fashion. In the next section, we will present our computer algebra representation and give the final result of the angular discretization in section~\ref{sec:real_valued_pn_eq}.

>>>>

Our method derives from the radiative transfer equation (RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
}\nonumber\\
&
+Q\left(\vec{x}, \omega\right)\nonumber
\label{eq:rte}
\end{align}

The left hand side (LHS) is the transport term, and we refer to the terms on the right hand side (RHS) as collision, scattering, and source term, respectively. The symbols $\sigma_t$, $\sigma_s$, $p$, and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term.

The RTE is often given in operator notation, where transport, collision, and scattering are expressed as operators $\mathcal{T}$, $\mathcal{C}$ and $\mathcal{S}$, which are applied to the radiance field $L$:
\begin{align}
\mathcal{T}\left(L\right) = -\mathcal{C}\left(L\right) + \mathcal{S}\left(L\right) + Q
\end{align}

Deterministic methods, like ours, are derived by discretizing the angular and spatial domain. This gives rise to a linear system of equations, which can be solved using standard methods. For the $P_N$-method, the angular variable is first discretized, using a truncated spherical harmonics expansion. This results in the $P_N$-equations, a system of coupled PDE's that still depend on a continuous spatial variable.

The number of equations grows with the truncation order $N$. This is why the discretization is hard and error prone if done by hand. We therefore use a computer algebra representation to automate this process. After giving an outline of the general discretization in this section, we will present our computer algebra representation in the next section. The $P_N$-equations that result from our automated discretization are given in Section~\ref{sec:real_valued_pn_eq}.

Since the radiance field $L$ is real, we use the real-valued SH basis functions $\SHBR^{l,m}$, which are defined in terms of the complex-valued SH basis functions $\SHBC^{l,m}$:
\begin{align}
\SHBR^{l,m}=
\left\{
\begin{array}{lr}
\frac{\iu}{\sqrt{2}}\left(\SHBC^{l,m}-\left(-1\right)^m\SHBC^{l,-m}\right), & \text{for } m < 0\\
\SHBC^{l,m}, & \text{for } m = 0\\
\frac{1}{\sqrt{2}}\left(\SHBC^{l,-m}-\left(-1\right)^m\SHBC^{l,m}\right), & \text{for } m > 0
\end{array}
\right.
\label{eq:sh_real_basis}
\end{align}

We express the projection into the spherical harmonics basis functions with a projection operator $\mathcal{P}$:
\begin{align}
\mathcal{P}^{l, m}(f) = \int_{\Omega}f(\mathbf{x}, \omega) \SHBR^{l,m}(\omega)\,\mathrm{d}\omega = f^{l,m}
\nonumber
\end{align}

The $P_N$-equations are derived by first expressing all direction-dependent parameters in spherical harmonics. The radiance field $L$ in Equation~\ref{eq:rte} is therefore replaced by its SH reconstruction $\widehat{L}$, introducing an error due to truncation at order $N$:
\begin{align}
\widehat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
}
\approx
L\left(\vec{x}, \omega\right)
\nonumber
\end{align}

After substitution, all angular parameters are expressed in terms of spherical harmonics, but they still depend on the continuous angular variable $\omega$. As a next step, we project each term of the RTE into spherical harmonics, using the projection operator $\mathcal{P}$. This produces a single equation for each $l,m$-pair. The $P_N$-equations therefore can be written as:
\begin{align}
\mathcal{P}^{l,m}\mathcal{T}\left(\widehat{L}\right)
=
-\mathcal{P}^{l,m}\mathcal{C}\left(\widehat{L}\right) 
+\mathcal{P}^{l,m}\mathcal{S}\left(\widehat{L}\right)
+\mathcal{P}^{l,m}\left(Q\right)
\label{eq:pn_operator_notation}
\end{align}

Once the $P_N$-equation have been found, the spatial variable $\vec{x}$ is discretized using a finite difference (FD) voxel grid (using central differences for differential operators).

Following this discretization, the radiance field $L$, is represented as a set of SH coefficients per voxel. Flattening these over all voxels into a single vector, gives the solution vector $\vec{u}$. The RHS vector $\vec{Q}$ is produced similarly. The projected operators can be expressed as linear transformations, which can be collapsed into a single coefficient matrix $A$ (see figure~\ref{fig:matrix_layout}):
\begin{align}
(T+C-S)\vec{u} = A\vec{u} = \vec{Q}
\end{align}

@@

So far, we have only given the $P_N$-equations in high-level operator notation (Equation~\ref{eq:pn_operator_notation}). Carrying out the full derivation creates large, unwieldy equations and requires a string of expansions, manipulations and applications of identities. These are challenging and error-prone if done by hand. We therefore developed a computer algebra representation, which allowed us to derive the the $P_N$-equation in a semi-automatic fashion.

### 2018-04-10 latex/egsr17-cgf/section_results.tex
final version

Classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
\begin{align} 
\nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
\label{eq:cda}
\end{align}

Since our solver can work with any PDE, which results in a linear system of equations, we only needed to put equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.

@@

First we run our solver on the 2D checkerboard, a very common test case in other fields. This allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}.

Their solver solves for the time-dependent and complex-valued $P_N$-equations in the 2D case, which is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent. This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish. Due to the time-dependency, their approach is to do explicit incremental steps in time. We run their solver for many timesteps, in order to get a result which is close to steady state.

\begin{figure}[h]
\centering
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p5_neumann_staggered_starmap.png}
\end{subfigure}%
\hspace{0.01\columnwidth}
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p5_neumann_staggered.png}
\end{subfigure}%
\vspace{-0.1in}
\icaption{Comparison of the result for the checkerboard test using Starmap~\cite{Seibold14} (left) against ours (right). Our result is in excellent agreement with the reference. We attribute the difference at the boundary to the lack of convergence of the Starmap solve.}
\label{fig:vs_starmap}
\end{figure}

As can be seen in figure~\ref{fig:vs_starmap}, the results from our solver are in excellent agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation.

@@

We also run our solver for the point source problem, a single point light in a homogeneous medium. This not only helps to validate our implementation for the 3D case, but also allows to get a clear assessment of the methods accuracy. We use the Grosjean approximation, which was introduced by D'Eon et al.\cite{dEon11}, as a very accurate approximation to the ground truth solution.

@@

For our test case, we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$. We run the solver for different truncation values $N$. In figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order. The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even. We further see, that the $P_1$ solution exactly matches the results from CDA, which confirmes, that the latter is only a collapsed version of the former.

\subsection{Procedural Clouds}

Finally, we run our solver on different procedural cloud datasets to get an idea of its performance in more practical applications.

Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function. We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected.

\begin{itemize}
	\item mention performance.
	\item run for various values of $N$
	\item convergence behaviour in vacuum. Clearify
	\item run for anisotropic problem
\end{itemize}

>>>>

Our computer algebra framework and the precomputation has been implemented in Python. The runtime component of our solver has been implemented in C++. We use a naive implementation of a CG solver, which has been modified such that we do not need to expliticly compute the normal form of the linear system to solve. We use the sparse matrix representation and sparse matrix vector product from the Eigen linear algebra library.

The solver for classical diffusion is based on the diffusion equation, which is derived by collapsing the $P_1$-equations:
\begin{align} 
\nabla\left(\frac{1}{3\sigma_t}\nabla L^{0,0}\right)  = -Q^{0,0}
\label{eq:cda}
\end{align}

Since our solver can work with any PDE, which results in a linear system of equations, we put Equation~\ref{eq:cda} into our computer algebra representation and provide it as an input to our solver, which would generate the correct stencil code automatically.

@@

First we ran our solver on the 2D checkerboard, a very common test case in other fields. The problem has dimensions $7\times 7$ and is discretized with resolution $71\times 71$. Unit size blocks are filled with purely absorbing medium with $\sigma_a=10$ in a checkerboard pattern. All other blocks are filled with a purely scattering medium with $\sigma_s=1$.

Solving the standard checkerboard problem, allows us to validate our solver against the solver from Seibold et al.~\cite{Seibold14}, which solves for the time-dependent and complex-valued $P_N$-equations in the 2D case only. The 2D case is derived by assuming that all SH coefficients, RTE parameters and boundary conditions are z-independent. This causes all SH coefficients and moment equations for which $l+m$ is odd to vanish. Due to the time-dependency, their approach is to do explicit incremental steps in time. We run their solver for many timesteps, in order to get a result which is close to steady state.

\begin{figure}[h]
\centering
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered_starmap.png}
\end{subfigure}%
\hspace{0.01\columnwidth}
\begin{subfigure}{0.49\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_neumann_staggered.png}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Comparison of the result for the checkerboard test using Starmaps time-stepping solver~\cite{Seibold14} (left) against our steady-state solver (right) with $N=5$. Our results are in good agreement.}
\label{fig:vs_starmap}
\end{figure}

As can be seen in figure~\ref{fig:vs_starmap}, the results from our solver are in good agreement with the results from Seibold et al.~\cite{Seibold14} and verify the correctness of our implementation. Converging to a residual of $10e^{-10}$ takes $0.27s$ for $P_1$ and $25s$ for $P_5$.

@@

We also run our solver for the point source problem, a single point light in a homogeneous medium. This not only helps to validate our implementation for the 3D case, but also provides information on the accuracy of these methods. We use the Grosjean approximation, which was introduced by D'Eon et al.\cite{dEon11} as a very accurate approximation to the ground truth solution.

@@

For our test case, we choose a FD resolution of $80\times80\times80$, an extinction coefficient $\sigma_t=8.0$ and albedo $\alpha=0.9$. We run the solver for different truncation values $N$. In Figure~\ref{fig:pointsource}~\subref{fig:pointsource_pn}, we see that the solution becomes increasingly accurate for higher truncation order. The ground truth is underestimated when $N$ is odd, and overestimated when $N$ is even. We further see, that the $P_1$ solution exactly matches the results from CDA, which confirmes, that the latter is only a collapsed version of the former. The time to solve is significant with $10m$ for $P_1$ and $45m$ with $P_5$. With these performance characteristics, the $P_N$-method is clearly not competitive in comparison with much diffusion and FLD solver, which are much faster.

\subsection{Nebulae}

Finally, we run our solver on a procedural cloud dataset to get an idea of its performance in more practical applications. Figure~\ref{fig:teaser}~\subref{fig:nebulae_ours} shows the result of $P_5$ for a procedurally generated heterogeneous cloud with an isotropic phase function. We see that at order $N=5$, our method can capture indirect illumination similarily well as FLD and is significantly better than CDA as expected. The indirectly illuminated region at the bottom appears to be closer to the path-traced result as opposed to the solution form FLD which is very flat in that region. However, in many other areas, $P_5$ seems to still suffer a lot from the problem of energy loss near transitions from dense regions to vacuum. It appears that going higher order a few steps, mitigates this problem only slowly at an immense cost of compute and storage.

The main characteristic of the nebulae dataset is the presence of vacuum. We found that having vacuum regions in the dataset will cause the condition number to become infinite and the solver practically doesnt converge. We therefore introduced a minimum threshold for the extinction coefficient $\sigma_t$. Every voxel with an extinction coefficient smaller than the threshold would be set to the threshold value. In Figure~\ref{fig:results_convergence} we show the effect of the minimum threshold on the convergence behavior. As the threshold increases, convergence improves.

%\begin{table}[h]
%\begin{center}
%\begin{tabular}{ c | c c c }
% & \textbf{2D Checkerboard} & \textbf{Point Source} & \textbf{Nebulae} \\ 
%\hline
%$P_1$ & 0.27s & 10m (64\times 64\times 64) & cell6 \\  
%$P_2$ & 2.4s & cell8 & cell9 \\
%$P_3$ & 5.8s & cell8 & cell9 \\
%$P_4$ & 12.7s & cell8 & cell9 \\
%$P_5$ & 25s & ? & ?
%\end{tabular}
%\end{center}
%\icaption{Performance measurements for different datasets and varying truncation order $N$}
%\label{table2}
%\end{table}

\begin{figure}[h]
\centering
%\missingfigure{test}
\includegraphics[width=0.9\columnwidth]{figures/fig_nebulae_p1_convergence.pdf}
%\vspace{-0.2in}
\vspace{-0.1in}
\icaption{Convergence behavior of our solver with $N=1$ for the nebulae dataset and for varying minimum thresholds of the extinction coefficient $\sigma_t$. Threshold values and an estimate for the condition number of $A$ (Matlabs condest function) are shown next to the plots. The convergence deteriorates as the threshold decreases. Once it reaches zero, the presence of pure vaccuum makes the condition number infinite.}
\label{fig:results_convergence}
\end{figure}

%\begin{itemize}
%	\item mention performance.
%	\item run for various values of $N$
%	\item convergence behaviour in vacuum. Clearify
%	\item run for anisotropic problem
%\end{itemize}

### 2018-04-10 latex/egsr17-cgf/solver.tex
final version

The truncation order $N$ is the key input parameter to the solver. With higher values, manually coding the solver from the equations would be ardous, error prone and very time consuming. We therefore decided to make use of the equations computer algebra representation and designed our solver around it.

The solver consists of two components. The first is a precomputation (section~\ref{sec:solver_precomputation}), which is done once for every single value of $N$. This steps runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space. The precomputation step automatically generates source code from the resulting expression tree.

The generated stencil code is compiled with the runtime component (section~\ref{sec:solver_runtime}) of our solver. This component gets the actual problem as input, including gid resolution and RTE parameter fields. It builds the linear system and solves it using standard methods. An overview of the solver is given in figure~\ref{fig:pnsolver}.

@@

The result of the precomputation is a stencil, which can be used during runtime to build the linear system for a given problem. The stencil is a pattern of indices into the solution vector, along with some weights. It expresses how the sum of the weighted solution vector components relate to the RHS for a given unknown in the system and therefore contains most information required to propagate a row in the system matrix $A$ and RHS-vector $\vec{Q}$. The index pattern does not change for different rows, giving the stencil its name.

A stencil generally is crafted by discretizing a PDE at a hypothetical center voxel $ijk$ (assuming to mean the voxel center most of the time). Finite differences create weighted references to other voxels (e.g. $i+1,jk$). After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and $ijk$-offsets. $ijk$ will only be known during runtime, when the stencil is executed for a particular unknown (row). Then the offsets can be used to find the component index into the solution-vector and weights can be evaluated for concrete world space position. We refer to the space with the hypthetical voxel $ijk$ at the center as stencil space.

Typically, the stencil is coded by hand, which would be challenging with our $P_N$-equations and higher values of $N$. In our system, the stencil code is created automatically from the computer algebra representation. Spatial discretization and fatorization into canonical form are done as manipulation passes on the mathematical expression tree. The precomputation code iterates over all terms and analyzes the unkown to get the $ijk$-offset. The factor-expression is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).

The spatial discretization is done by parsing the expression tree of each equation from the root. All occurances of the continuous positional variable $\vec{x}$ are replaced by discrete voxel coordinates $i,j,k$, whose values are retrieved from the top of a stack. The stack is kept by the parser and is initialized with the voxel at $ijk=(0,0,0)$.

Whenever the parser encounters a differential operator, the current voxel position is shifted and pushed to the stack. Then the child expression is parsed and multiplied with a weight value, followed by popping the stack. This is done for each element of a central difference stencil, which provides the offset and weight values (referencing the voxelsize variable). The dimension of the stencil is given by the derivation variable of the operator. Nested derivatives are handled naturally and produce higher order finite difference stencils as expected.

Another aspect to consider during discretization is the location of the variables and unknowns. As we will see later, our solver needs to support staggered grids, where the SH coefficients are not all located at the voxel center (collocated). Coefficients can also be located at face-centers, edges and intersections of voxels (staggered).

Our solver supports placement of coefficients at arbitrary staggered grid locations. Whenever a coefficient is encountered during expression tree parsing for discretization, the parser checks how its location is related to the current location on the top of the stack. Depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at the requested offset from its surrounding locations, or it returns the coefficient itself, if it happens to coincide. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center. The position stack of the parser is initialized with the staggered grid location defined by the coefficient, which is associated with the $l,m$-pair of the equation currently being discretized.

@@

The generated stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver. The runtime executes the stencil for every voxel to propagate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.

The number of rows is determined by the number of voxels times the number of coefficients per voxel (see figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order. The matrix $A$ is squared and fortunately very sparse, due to finite differences and the structure of the $P_N$-equations. Unfortunately, it also is non-symmetric (due to the transport term) and not diagonal dominant, which rules out many standard methods for solving linear systems. We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.

The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC). Our solver supports Neumann BC and Dirichlet BC. They are handled transparently by the code which runs the stencil. Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or write into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC).

In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}). This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the runtime code.

\begin{figure}[h]
\centering
\includegraphics[width=0.5\columnwidth]{figures/fig_staggered_grids.pdf}
%\vspace{-0.2in}
\icaption{Staggered grids place coefficients at different locations within the finite difference grid. Additional unknowns are required in boundary cells to produce a correct result.}
\label{fig:staggeredgrid}
\end{figure}

>>>>

The truncation order $N$ is the key input parameter to the solver. With higher values, manual implementation of the solver from the equations would be ardous, error-prone and time-consuming. We therefore decided to make use of the computer algebra representation and designed our solver around it.

The solver consists of two components. The first is a precomputation (Section~\ref{sec:solver_precomputation}), which is excuted once for every single value of $N$. This steps runs a partial evaluation on the mathematical expression tree and applies the spatial discretization in a reference space we call stencil space. The precomputation step automatically generates source code from the resulting expression tree.

The generated stencil code is compiled with the runtime component (Section~\ref{sec:solver_runtime}) of our solver. This component receives the actual problem as input, including grid resolution and RTE parameter fields. It then builds the linear system and solves it using standard methods. An overview of the solver is given in Figure~\ref{fig:pnsolver}.

@@

The result of the precomputation is a stencil, which can be used during runtime to build the linear system for a given problem. The stencil is a pattern of indices into the solution vector, along with values. It expresses how the sum of the weighted solution vector, components relate to the RHS for a given unknown in the system and therefore contains most information required to fill the system matrix $A$ and RHS-vector $\vec{Q}$ row by row. Note that while the coefficients may change, the sparsity pattern of the stencil is identical for different rows.

\begin{figure}[b]
\centering
%\missingfigure{test}
\includegraphics[width=0.7\columnwidth]{figures/fig_stencil_pipeline.pdf}
%\vspace{-0.2in}
\icaption{Creating the stencil code requires several steps, usually done by hand. We express the given problem in a computer algebra representation and use it to fully automate the process.}
\label{fig:stencile_pipeline}
\end{figure}

Stencil generation entails discretizing a PDE at a hypothetical center voxel $(i, j, k)$ (assuming to mean the voxel center most of the time). Finite differences create weighted references to other voxels (e.g. $i+1, j, k$). After bringing the discretized equation into canonical form (a weighted sum of unknowns), one can write the stencil by reading off the weights and offsets (Figure~\ref{fig:stencile_pipeline}). $(i, j, k)$ will only be known during runtime, when the stencil is executed for a particular unknown (row). Then the offsets can be used to find the component index into the solution-vector, and weights can be evaluated for concrete world space position. We refer to the space with the hypthetical voxel $(i, j, k)$ at the center as stencil space.

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{figures/fig_discretization.pdf}
%\end{center}
\end{wrapfigure}The spatial discretization is done by parsing the expression tree from the root. The discrete substitute for the continuous position variable $\vec{x}$ is initialized with $ijk$. Differential operator nodes are replaced by a subtree, which expresses the finite difference approximation (including voxelsize factor $h$). The subtree of the differential operator node is duplicated for different offsets to the discrete variable $(i, j, k)$. Since this offset only applies to the substree, a stack is maintained by the parser to manage scope. Whenever the parser encounters the continuous variable $\vec{x}$ in the expression tree, its node in the tree is replaced by the discrete substitute, currently on top of the stack. Nested differential operators yield higher order finite difference stencils as expected.

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.2in}
\includegraphics[width=0.6\columnwidth]{figures/fig_codegen.pdf}
%\end{center}
\end{wrapfigure}
Factorization into canonical form is done as a manipulation pass on the mathematical expression tree. The result allows to implement the code generation in a straight forward fashion. For each term, the $ijk$-offset is retrieved from the unknown. The factor-expression, which is multipled with the unknown, is extracted from the tree and used to generate source code for evaluating the factor-expression during runtime (including calls for evaluating RTE-parameter fields).

\begin{figure}[h]

%\vspace{1in}
\centering
\includegraphics[width=0.9\columnwidth]{figures/fig_staggered_grids.pdf}
%\vspace{-0.2in}
\icaption{Staggered grids place coefficients at different locations within the finite difference grid.}
\label{fig:staggeredgrid}
\end{figure}

\begin{wrapfigure}{r}{0.6\columnwidth}
%\begin{center}
\hspace{-.1in}
\includegraphics[width=0.6\columnwidth]{figures/fig_staggered_interpolation.pdf}
\vspace{-.1in}
%\end{center}
\end{wrapfigure}Our solver supports placement of coefficients at arbitrary staggered grid locations (see Figure~\ref{fig:staggeredgrid}). This means, that during the discretization step, the discrete location $(i, j, k)$ (at which the unknown is meant to be evaluated) might not coincide with the location of the unknown. To solve this, depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at $(i, j, k)$ from their defined locations. It those happen to coincide, the coefficient itsself is returned. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center.

@@

The stencil code is generated once for every value of $N$ and compiled with the runtime component of our solver. The runtime executes the stencil for every voxel to propagate the system matrix $A$ and RHS vector $\vec{Q}$ with numerical values.

The number of rows is determined by the number of voxels times the number of coefficients per voxel (see Figure~\ref{fig:matrix_layout}) and can therefore become very large for high resolution and high truncation order. The matrix $A$ is square and extremely sparse, due to the local structure of the finite differences discretization. Unfortunately, it is non-symmetric due to the transport term and not diagonal dominant, which rules out many standard methods for solving linear systems. We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number. Investigation of other solution schemes (e.g. multigrid) would be an interesting avenue for future work.

Our solver supports Neumann and Dirichlet boundary conditions. They are handled transparently by the code which generates the stencil. Whenever the stencil accumulates coefficients into a boundary location, the framework either ignores the write operation (Dirichlet BC) or accumulates into the row and column in $A$ of the coefficient in the closest voxel inside the domain (Neumann BC). This is done changing the index of the written component under the hood.

%To respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels. This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the runtime code.
